---
title: "Assignment 1: Quarto template"
format: gfm
---

## Load packages

First, load all the required packages with the following code. Install them if they are not installed yet.


```{r label00, echo=TRUE, results='asis'}
library(tidyverse)
library(knitr)
library(gapminder)
```

## Exercise 1. Trends in land value

This excercise uses a dataset that describes the trends in land value (```Land.Value```), among other variables, in different states in the US 1975-2013. The states are grouped into four different regions, under the variable region. This dataset was obtained from the Data Science Services of Harvard University.

```{r label0, echo=TRUE, results='asis'}
housing <- read_csv("https://raw.githubusercontent.com/nt246/NTRES-6100-data-science/master/datasets/landdata_states.csv")
housing %>%
  head() %>% 
  kable() 
```

#### 1.1 1.1 Washington DC was not assigned to a region in this dataset. According to the United States Census Bureau, however, DC is part of the South region. Here:

* Change the region of DC to “South” (Hint: there are multiple ways to do this, but ```mutate()``` and ```ifelse()``` might be helpful)

* Create a new tibble or regular dataframe consisting of this new updated region variable along with the original variables State, Date and Land.Value (and no others)

* Pull out the records from DC in this new data frame. How many records are there from DC? Show the first 6 lines.

```{r label1, echo=TRUE, results='asis'}

housing2 <- housing %>%
  mutate(region = ifelse(State == "DC", "South", region)) %>%
  select(region, State, Land.Value, Date)

dc_data <- housing2 %>% filter(State == "DC")

n_dc <- nrow(dc_data)

n_dc
head(dc_data)
```


There are 153 records for DC.


```


#### 1.2 Generate a tibble/dataframe that summarizes the mean land value of each region at each time point and show its first 6 lines.

```{r label2, echo=TRUE, fig.width=6, fig.height=4}
mean_land <- housing2 %>%
  group_by(region, Date) %>%
  summarize(mean_land_value = mean(Land.Value, na.rm = TRUE), .groups = "drop")

head(mean_land)
```

```

#### 1.3 Using the tibble/dataframe from 1.2, plot the trend in mean land value of each region through time

```{r label3, echo=TRUE, fig.width=6, fig.height=4}
ggplot(mean_land, aes(x = Date, y = mean_land_value, color = region)) +
  geom_line(size = 0.5) +
  labs(title = "Trends in Mean Land Value by Region (1975–2013)",
       x = "Year",
       y = "Mean Land Value (USD)",
       color = "Region") +
  theme_gray()
```

<br>

## Exercise 2. Life expectancy and GDP per capita 1952-2007

This exercise uses the ```gapminder``` dataset from the gapminder package. It describes the life expectancy (```lifeExp```), GDP per capita (```gdpPercap```), and population (```pop```) of 142 countries from 1952 to 2007. These countries can be grouped into 5 continents. As a reminder, **reproduce the following plots exactly as shown**.

Have a look at the data structure

```{r label13, echo=TRUE, results='asis'}
gapminder %>% 
  head() %>% 
  kable()
```

#### 2.1 Use a scatterplot to explore the relationship between per capita GDP (gdpPercap) and life expectancy (lifeExp) in the year 2007.

```{r filter-dose-time, echo=TRUE, results='asis'}

gapminder_2007 <- gapminder %>% filter(year == 2007)

ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  labs(title = "Life Expectancy vs GDP per Capita (2007)",
       x = "GDP per Capita (USD)",
       y = "Life Expectancy (Years)") +
  theme_gray()

```

#### 2.2 Add a smoothing line to the previous plot.

```{r label14, echo=TRUE, results='asis'}
gapminder_2007 <- gapminder %>% filter(year == 2007)

ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Life Expectancy vs GDP per Capita (2007)",
       x = "GDP per Capita (USD)",
       y = "Life Expectancy (Years)") +
  theme_gray()
```

#### 2.3 Exclude Oceania from the previous plot, show each continent in a different color, and fit a separate smoothing line to each continent to identify differences in this relationship between continents. Turn off the confidence intervals.
Note: only two Oceanian countries are included in this dataset, and ```geom_smooth()``` does not work with two data points, which is why they are excluded.


```{r label15, echo=TRUE, results='asis'}
gapminder_2007 <- gapminder %>%
  filter(year == 2007, continent != "Oceania")

ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(
    title = "Life Expectancy vs GDP per Capita (2007, excluding Oceania)",
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (Years)",
    color = "Continent"
  ) +
  theme_gray()
```

#### 2.4 Use faceting to solve the same problem. Include the confidence intervals in this plot.

```{r label16, echo=TRUE, results='asis'}
gapminder_2007 <- gapminder %>%
  filter(year == 2007, continent != "Oceania")

ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  geom_smooth(se = TRUE) +
  facet_wrap(~ continent) +
  labs(
    title = "Life Expectancy vs GDP per Capita (2007, excluding Oceania)",
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (Years)"
  ) +
  theme_gray()
```

#### 2.5 Explore the trend in life expectancy through time in each continent. Color by continent.

```{r label17, echo=TRUE, results='asis'}
ggplot(gapminder, aes(x = year, y = lifeExp, color = continent)) +
  geom_line(aes(group = country), alpha = 0.5) +
  facet_wrap(~ continent) +
  labs(
    title = "Life Expectancy Trends Over Time by Continent",
    x = "Year",
    y = "Life Expectancy (Years)"
  ) +
  theme_gray()

```
#### 2.6 From the previous plot, we see some abnormal trends in Asia and Africa, where the the life expectancy in some countries sharply dropped at certain time periods. Here, we look into what happened in Asia in more detail. First, create a new dataset by filtering only the Asian countries. Show the first 6 lines of this filtered dataset.


```{r label18, echo=TRUE, results='asis'}
gapminder_asia <- gapminder %>%
  filter(continent == "Asia")
head(gapminder_asia)
```
<br>

### 2.7 Using the filtered dataset, identify the countries that had abnormal trends in life expectancy by plotting, and discuss historical events possibly explaining these trends. (Hint: facet by country)

```{r label19, echo=TRUE, results='asis'}
ggplot(gapminder_asia, aes(x = year, y = lifeExp)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ country) +
  labs(
    title = "Life Expectancy Trends in Asian Countries (1952–2007)",
    x = "Year",
    y = "Life Expectancy (Years)"
  ) +
  theme_gray()
```

Abnormal trends: Cambodia (Cambodian civil war and genocide), China (population density), Iraq (civil conflicts), Korea Dem Rep (fascism).



